---
title: "NCAA March Madness Bracket Models — Student Assignment"
subtitle: "STA345 — Spring 2026"
author: "Alejandro Alfaro and Alex Barroso"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

In this assignment you will:

1. **Understand** the Power Model from Ludden et al. (2020) and prove its circularity
2. **Implement** the Power Model from scratch in R
3. **Build** your own improved probability model
4. **Evaluate** your model against the Power Model using bracket generators

**Estimated time:** 1.5 -- 2 hours

**Data provided:**

- `data/combined.csv` — NCAA tournament game results (1985--2024)
- `data/metrics.csv` — Team-level adjusted efficiency metrics (2002--2026)

---

# Part 1: Understanding the Power Model (30 min)

## 1.1 Load and explore the data

```{r}
library(tidyverse)

games <- read_csv("data/combined.csv", show_col_types = FALSE) %>%
  mutate(round = case_when(
    round_of == 64 ~ 1, round_of == 32 ~ 2, round_of == 16 ~ 3,
    round_of == 8  ~ 4, round_of == 4  ~ 5, round_of == 2  ~ 6
  ))

# Look at the data
glimpse(games)
```

**Q1.1:** How many total games are in the dataset? How many tournament years? What columns are available?

```{r}
# YOUR CODE HERE
```

## 1.2 Compute Round 1 win rates

In March Madness Round 1, the matchups are always fixed by seed: 1 vs 16, 2 vs 15, 3 vs 14, 4 vs 13, 5 vs 12, 6 vs 11, 7 vs 10, 8 vs 9.

```{r}
r1_stats <- games %>%
  filter(round == 1) %>%
  mutate(
    s1 = pmin(winning_team_seed, losing_team_seed),  # better seed
    s2 = pmax(winning_team_seed, losing_team_seed),  # worse seed
    s1_won = (winning_team_seed == s1)                # did better seed win?
  ) %>%
  group_by(s1, s2) %>%
  summarise(
    games = n(),
    wins = sum(s1_won),
    p_bar = mean(s1_won),
    .groups = "drop"
  )

r1_stats
```

**Q1.2:** Which Round 1 matchup has the closest win rate to 50/50? Which is the most lopsided? Do any results surprise you?

## 1.3 The Power Model formula

The Power Model predicts:

$$P(s_1 \text{ beats } s_2) = \frac{s_2^{\alpha}}{s_1^{\alpha} + s_2^{\alpha}}$$

where $\alpha$ is a parameter. For Round 1, $\alpha$ is computed separately for each matchup by inverting the formula:

$$\alpha = \frac{\text{logit}(\bar{p})}{\log(1/s_1) - \log(1/s_2)}$$

where $\bar{p}$ is the observed win rate and $\text{logit}(p) = \log(p / (1-p))$.

**Q1.3 (Derivation):** Starting from the Power Model formula, show algebraically that solving for $\alpha$ from $\bar{p}$ and then plugging it back into the formula returns $\bar{p}$ exactly. In other words, prove that the Power Model is circular for Round 1.

*Hint:* Start by writing $\bar{p} = s_2^{\alpha} / (s_1^{\alpha} + s_2^{\alpha})$, take the logit of both sides, solve for $\alpha$, then substitute back.

**YOUR DERIVATION HERE** (write it out by hand or in LaTeX)

## 1.4 Verify computationally

Now verify your derivation with code:

```{r}
compute_alpha <- function(p_bar, s1, s2) {
  logit_p <- log(p_bar / (1 - p_bar))
  alpha <- logit_p / (log(1/s1) - log(1/s2))
  return(alpha)
}

power_model_prob <- function(alpha, s1, s2) {
  s2^alpha / (s1^alpha + s2^alpha)
}

# Compute alpha and p_hat for each R1 matchup
r1_check <- r1_stats %>%
  rowwise() %>%
  mutate(
    alpha = compute_alpha(p_bar, s1, s2),
    p_hat = power_model_prob(alpha, s1, s2),
    difference = abs(p_bar - p_hat)
  ) %>%
  ungroup()

r1_check %>% select(s1, s2, p_bar, alpha, p_hat, difference)
```

**Q1.4:** What is the largest value in the `difference` column? What does this tell you about the Power Model?

---

# Part 2: Implement the Power Model (30 min)

## 2.1 Build the full probability matrix

The Power Model produces a 16 x 16 x 6 array: `prob_mat[s1, s2, round]` gives the probability that seed `s1` beats seed `s2` in a given round.

- **Round 1:** Use matchup-specific alphas (computed above)
- **Rounds 2--6:** Compute a single pooled alpha per round from all games in that round

Complete the function below:

```{r}
build_power_model <- function(data, train_years) {
  train <- data %>%
    filter(year %in% train_years) %>%
    mutate(
      s1 = pmin(winning_team_seed, losing_team_seed),
      s2 = pmax(winning_team_seed, losing_team_seed),
      s1_won = (winning_team_seed == s1)
    )

  # Round 1: matchup-specific alphas
  r1_alphas <- train %>%
    filter(round == 1) %>%
    group_by(s1, s2) %>%
    summarise(p_bar = mean(s1_won), n = n(), .groups = "drop") %>%
    rowwise() %>%
    mutate(alpha = compute_alpha(p_bar, s1, s2)) %>%
    ungroup()

  # Rounds 2-6: pooled alpha per round
  # TODO: Complete this section
  # 1. Filter training data for rounds >= 2 where s1 != s2
  # 2. For each (round, s1, s2) combination, compute p_bar and alpha
  # 3. For each round, compute the weighted mean of alpha (weighted by n)
  round_alphas <- train %>%
    filter(round >= 2, s1 != s2) %>%
    # YOUR CODE HERE: group, summarise, compute alphas, then get weighted mean per round
    group_by(round, s1, s2) %>%
    summarise(p_bar = mean(s1_won), n = n(), .groups = "drop") %>%
    rowwise() %>%
    mutate(alpha = compute_alpha(p_bar, s1, s2)) %>%
    ungroup() %>%
    group_by(round) %>%
    summarise(alpha_bar = weighted.mean(alpha, w = n), .groups = "drop")

  list(r1_alphas = r1_alphas, round_alphas = round_alphas)
}
```

## 2.2 Build the probability matrix

```{r}
build_prob_matrix <- function(model) {
  mat <- array(0.5, dim = c(16, 16, 6))

  # Look up round-level alphas
  round_alpha_lookup <- rep(1, 6)
  for (i in seq_len(nrow(model$round_alphas))) {
    round_alpha_lookup[model$round_alphas$round[i]] <- model$round_alphas$alpha_bar[i]
  }

  for (s1 in 1:15) {
    for (s2 in (s1+1):16) {
      # Round 1: matchup-specific alpha
      row <- model$r1_alphas %>% filter(.data$s1 == !!s1, .data$s2 == !!s2)
      if (nrow(row) > 0) {
        p <- power_model_prob(row$alpha[1], s1, s2)
      } else {
        p <- power_model_prob(1.0, s1, s2)  # default
      }
      mat[s1, s2, 1] <- p
      mat[s2, s1, 1] <- 1 - p

      # Rounds 2-6: round-level alpha
      for (rnd in 2:6) {
        p <- power_model_prob(round_alpha_lookup[rnd], s1, s2)
        mat[s1, s2, rnd] <- p
        mat[s2, s1, rnd] <- 1 - p
      }
    }
  }
  mat
}

# Test it
train_years <- setdiff(1985:2023, 2020)
model <- build_power_model(games, train_years)
prob_mat <- build_prob_matrix(model)

# Check: P(1-seed beats 16-seed in R1)
cat("P(1 beats 16, R1):", round(prob_mat[1, 16, 1], 4), "\n")
cat("P(8 beats 9, R1):", round(prob_mat[8, 9, 1], 4), "\n")
cat("P(1 beats 2, R2):", round(prob_mat[1, 2, 2], 4), "\n")
```

**Q2.2:** Does `P(1 beats 16, R1)` match the observed frequency from Part 1? Why or why not?

---

# Part 3: Bracket Generation & Scoring (20 min)

## 3.1 The R64 Generator

The simplest generator: simulate every game using the probability matrix.

```{r}
# Standard NCAA bracket pairings
REGION_SEEDS <- matrix(c(1,16, 8,9, 5,12, 4,13, 6,11, 3,14, 7,10, 2,15),
                       ncol = 2, byrow = TRUE)

sim_game <- function(seed_a, seed_b, rnd, prob_mat) {
  p_a <- prob_mat[seed_a, seed_b, rnd]
  if (is.na(p_a)) p_a <- 0.5
  if (runif(1) < p_a) seed_a else seed_b
}

simulate_region <- function(prob_mat) {
  r1 <- integer(8)
  for (i in 1:8) r1[i] <- sim_game(REGION_SEEDS[i,1], REGION_SEEDS[i,2], 1L, prob_mat)
  r2 <- integer(4)
  for (i in 1:4) r2[i] <- sim_game(r1[2*i-1], r1[2*i], 2L, prob_mat)
  r3 <- c(sim_game(r2[1], r2[2], 3L, prob_mat), sim_game(r2[3], r2[4], 3L, prob_mat))
  r4 <- sim_game(r3[1], r3[2], 4L, prob_mat)
  list(r1 = r1, r2 = r2, r3 = r3, r4 = r4)
}

gen_R64 <- function(prob_mat) {
  reg <- lapply(1:4, function(i) simulate_region(prob_mat))
  f4 <- sapply(reg, function(r) r$r4)
  f4w <- c(sim_game(f4[1], f4[2], 5L, prob_mat),
           sim_game(f4[3], f4[4], 5L, prob_mat))
  champ <- sim_game(f4w[1], f4w[2], 6L, prob_mat)
  list(regions = reg, f4_seeds = f4, f4_winners = f4w, champion = champ)
}

# Generate one bracket
set.seed(42)
bracket <- gen_R64(prob_mat)
cat("Champion seed:", bracket$champion, "\n")
cat("Final Four seeds:", bracket$f4_seeds, "\n")
```

## 3.2 ESPN Scoring

```{r}
ESPN_POINTS <- c(10L, 20L, 40L, 80L, 160L, 320L)

precompute_actual <- function(data, target_year) {
  actual <- data %>% filter(year == target_year)
  round_of_map <- c(64, 32, 16, 8, 4, 2)
  lapply(1:6, function(rnd) {
    sort(actual %>% filter(round_of == round_of_map[rnd]) %>% pull(winning_team_seed))
  })
}

count_matches <- function(sim_seeds, actual_seeds) {
  sim_t <- tabulate(sim_seeds, nbins = 16)
  act_t <- tabulate(actual_seeds, nbins = 16)
  sum(pmin(sim_t, act_t))
}

score_bracket <- function(bracket, actual_rounds) {
  score <- 0L
  for (rnd in 1:3) {
    field <- c("r1", "r2", "r3")[rnd]
    sim <- unlist(lapply(bracket$regions, function(r) r[[field]]))
    score <- score + count_matches(sim, actual_rounds[[rnd]]) * ESPN_POINTS[rnd]
  }
  sim_r4 <- sapply(bracket$regions, function(r) r$r4)
  score <- score + count_matches(sim_r4, actual_rounds[[4]]) * ESPN_POINTS[4]
  score <- score + count_matches(bracket$f4_winners, actual_rounds[[5]]) * ESPN_POINTS[5]
  score <- score + count_matches(bracket$champion, actual_rounds[[6]]) * ESPN_POINTS[6]
  score
}

# Score the bracket against 2024 actual results
actual_2024 <- precompute_actual(games, 2024)
cat("Bracket ESPN score:", score_bracket(bracket, actual_2024), "/ 1920\n")
```

## 3.3 Generate a pool and find the max

```{r}
set.seed(123)
N <- 200
scores <- sapply(1:N, function(i) {
  b <- gen_R64(prob_mat)
  score_bracket(b, actual_2024)
})

cat("Pool of", N, "brackets:\n")
cat("  Max score:", max(scores), "\n")
cat("  Mean score:", round(mean(scores), 0), "\n")
cat("  Median score:", median(scores), "\n")
```

**Q3.3:** Run this with different values of N (100, 500, 1000). How does the max score change? Why?

---

# Part 4: Build Your Own Model (30 min)

Now it's your turn. Build a model that outperforms the Power Model.

## 4.1 Load team efficiency data

```{r}
metrics <- read_csv("data/metrics.csv", show_col_types = FALSE) %>%
  mutate(AdjEM = AdjOE - AdjDE)

# Join to game data
metrics_slim <- metrics %>%
  select(team = `Mapped ESPN Team Name`, Season, AdjEM)

games_rated <- games %>%
  left_join(metrics_slim, by = c("year" = "Season", "winning_team_name" = "team")) %>%
  rename(win_AdjEM = AdjEM) %>%
  left_join(metrics_slim, by = c("year" = "Season", "losing_team_name" = "team")) %>%
  rename(lose_AdjEM = AdjEM)

# Look at the relationship between AdjEM difference and winning
games_rated %>%
  filter(!is.na(win_AdjEM), !is.na(lose_AdjEM)) %>%
  mutate(em_diff = win_AdjEM - lose_AdjEM) %>%
  ggplot(aes(x = em_diff)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "AdjEM Difference (Winner - Loser)", y = "Count",
       title = "Winners tend to have higher AdjEM, but not always") +
  theme_minimal()
```

## 4.2 Build a probability matrix using AdjEM

**Your task:** Build a 16 x 16 x 6 probability matrix using the AdjEM data. Here's a skeleton:

```{r}
build_your_model <- function(data, train_years) {
  # Step 1: Filter training data to games with AdjEM
  train <- data %>%
    filter(year %in% train_years, !is.na(win_AdjEM), !is.na(lose_AdjEM)) %>%
    mutate(
      s1 = pmin(winning_team_seed, losing_team_seed),
      s2 = pmax(winning_team_seed, losing_team_seed),
      s1_won = as.integer(winning_team_seed == s1),
      em_diff = ifelse(s1_won == 1, win_AdjEM - lose_AdjEM, lose_AdjEM - win_AdjEM)
    ) %>%
    filter(s1 != s2)

  # Step 2: Fit a logistic regression
  # TODO: Try different formulas. Some ideas:
  #   - s1_won ~ em_diff
  #   - s1_won ~ em_diff + round
  #   - s1_won ~ em_diff * round + log(s2/s1)
  fit <- glm(s1_won ~ em_diff, data = train, family = binomial())

  # Step 3: Get average AdjEM by seed from training data
  seed_em <- bind_rows(
    train %>% select(seed = winning_team_seed, em = win_AdjEM),
    train %>% select(seed = losing_team_seed, em = lose_AdjEM)
  ) %>%
    group_by(seed) %>%
    summarise(mean_em = mean(em, na.rm = TRUE), .groups = "drop")

  # Step 4: Build probability matrix
  mat <- array(0.5, dim = c(16, 16, 6))
  for (s1 in 1:15) {
    for (s2 in (s1+1):16) {
      em1 <- seed_em$mean_em[seed_em$seed == s1]
      em2 <- seed_em$mean_em[seed_em$seed == s2]
      if (length(em1) == 0 || length(em2) == 0) next
      # TODO: Predict probability using your model
      p <- predict(fit, newdata = data.frame(em_diff = em1 - em2), type = "response")
      p <- max(0.01, min(0.99, p))
      for (rnd in 1:6) {
        mat[s1, s2, rnd] <- p
        mat[s2, s1, rnd] <- 1 - p
      }
    }
  }
  mat
}
```

## 4.3 Compare your model to the Power Model

```{r}
# Test on 2024
train_yrs <- setdiff(2002:2023, 2020)
train_yrs <- train_yrs[train_yrs %in% unique(games_rated$year)]

# Build both models
power_mat <- build_prob_matrix(build_power_model(games_rated, train_yrs))
your_mat <- build_your_model(games_rated, train_yrs)

actual_2024 <- precompute_actual(games_rated, 2024)

# Generate pools and compare
set.seed(456)
N <- 500

power_scores <- sapply(1:N, function(i) score_bracket(gen_R64(power_mat), actual_2024))
your_scores <- sapply(1:N, function(i) score_bracket(gen_R64(your_mat), actual_2024))

cat("=== 2024 Results (N=500 brackets) ===\n")
cat("Power Model -- Max:", max(power_scores), " Mean:", round(mean(power_scores), 0), "\n")
cat("Your Model  -- Max:", max(your_scores), " Mean:", round(mean(your_scores), 0), "\n")
cat("Improvement:", max(your_scores) - max(power_scores), "points\n")
```

**Q4.3:** Run this comparison for multiple years (2019, 2021, 2022, 2023, 2024). Does your model consistently beat the Power Model?

## 4.4 Challenge: Can you do better?

Ideas to try:

- Add `round` as a feature (or interact it with `em_diff`)
- Use `AdjOE` and `AdjDE` separately instead of just `AdjEM`
- Try a random forest instead of logistic regression
- Use year-specific AdjEM values instead of averages across all training years

```{r}
# YOUR IMPROVED MODEL HERE
```

---

# Part 5: Short Answer Questions (10 min)

**Q5.1:** In your own words, explain why the Power Model's alpha parameterization is circular for Round 1 matchups. Why doesn't this matter for Rounds 2--6?

**Q5.2:** The Power Model uses one alpha per round for Rounds 2--6. Give a specific example of why this is problematic (name two different seed matchups in the same round that should have very different win probabilities).

**Q5.3:** Why does using team-level AdjEM data break through the seed-only performance ceiling? What information does AdjEM capture that seed numbers alone cannot?

**Q5.4:** Our AdjEM model loses to the Power Model in 3 out of 11 test years (2014, 2018, 2022). All three were extremely upset-heavy tournaments. Why does the AdjEM model struggle in these years? What could you do to make the model more robust?

---

# Submission

Submit:

1. This `.Rmd` file with all code cells completed and questions answered
2. A short summary (3--5 sentences) of your model's performance vs the Power Model
