---
title: "March Madness Bracket Generation"
subtitle: "STA 345 — Student Assignment"
author: "Alex Barroso & Alejandro Alfaro"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Each year, millions of fans fill out NCAA tournament brackets trying to predict all 63 game winners. Ludden et al. (2020) propose a mathematical approach: instead of crafting one "perfect" bracket, generate a large *pool* of random brackets using a probability model, hoping at least one scores well.

In this assignment you'll explore the data behind their approach, run the simulation yourself, and tweak one parameter to see how it changes results.

## Setup

Load the shared project code and data. Make sure the `R/` and `data/` folders are in your working directory.

```{r load}
source("R/utils.R")
source("R/models.R")
source("R/generators.R")
source("R/scoring.R")

games <- read_csv("data/combined.csv") %>%
  mutate(round = case_when(
    round_of == 64 ~ 1, round_of == 32 ~ 2, round_of == 16 ~ 3,
    round_of ==  8 ~ 4, round_of ==  4 ~ 5, round_of ==  2 ~ 6
  ))

# Helper: evaluate a generator by running N brackets x R replications
evaluate_gen <- function(gen_fn, name, N, R, actual) {
  all_scores <- numeric(N * R)
  rep_maxes <- numeric(R)
  for (rep in 1:R) {
    offset <- (rep - 1) * N
    best <- 0
    for (i in 1:N) {
      s <- score_bracket_fast(gen_fn(), actual)
      all_scores[offset + i] <- s
      if (s > best) best <- s
    }
    rep_maxes[rep] <- best
  }
  tibble(generator = name,
         max_score_mean = mean(rep_maxes),
         max_score_sd = sd(rep_maxes),
         mean_score = mean(all_scores),
         scores = list(all_scores))
}
```

## Part 1: Understanding Upsets

The foundation of any bracket model is knowing how often upsets happen.

**Q1.** Run the code below. Which round has the highest upset rate? Why do you think that is?

```{r q1}
games %>%
  mutate(upset = winning_team_seed > losing_team_seed) %>%
  group_by(round) %>%
  summarise(games = n(), upsets = sum(upset), upset_rate = round(mean(upset), 3))
```

*Your answer:*

**Q2.** Now look at Round 1 matchups specifically. Which seed pairing is closest to a coin flip? Why do you think that is?

```{r q2}
games %>%
  filter(round == 1) %>%
  mutate(s1 = pmin(winning_team_seed, losing_team_seed),
         s2 = pmax(winning_team_seed, losing_team_seed),
         s1_won = winning_team_seed == s1) %>%
  group_by(matchup = paste0(s1, " vs ", s2)) %>%
  summarise(games = n(), stronger_seed_wins = mean(s1_won)) %>%
  arrange(stronger_seed_wins)
```

*Your answer:*

## Part 2: The Power Model's Key Idea

The paper models win probability as a function of the two seeds. The key parameter is **α (alpha)**: higher α means the stronger seed is more dominant; α near 0 means it's basically a coin flip.

```{r power-model}
power_mod <- build_power_model(games, setdiff(1985:2023, 2020))
```

**Q3.** Look at the round-level alpha values below. The paper argues that α varies meaningfully across rounds, meaning upsets aren't equally likely everywhere. Based on these values, in which round are upsets most common? In which round do favorites dominate most? Why do you think this is?

```{r q3}
power_mod$round_alphas
```

*Your answer:*

## Part 3: Simulate and Score Brackets

Now let's generate brackets and see how they score against a real tournament. The code below generates 500 brackets using two strategies and scores them against the 2024 tournament results:

-   **Pick Favorite:** Always pick the lower seed. No randomness.
-   **Power Model (R64):** Simulate every game using the power model probabilities.

```{r simulate}
prob_mat <- build_power_prob_matrix(power_mod)
actual <- precompute_actual(games, 2024)

set.seed(345)

pf_scores <- replicate(500, score_bracket_fast(gen_PF(), actual))
r64_scores <- replicate(500, score_bracket_fast(gen_R64(prob_mat), actual))

results <- tibble(
  score = c(pf_scores, r64_scores),
  method = rep(c("Pick Favorite", "Power Model"), each = 500)
)
```

**Q4.** Compare the two methods using any visualizations/summary statistics you see fit. How do they compare? Is there an argument for using the power model?

```{r q5}
# your code here
```

*Your answer:*

## Part 4: Tweak the Model 

The power model's α values control upset frequency. What if you think the model is **too conservative** (doesn't predict enough upsets) or **too aggressive**?

The code below lets you scale all the α values by a multiplier:

-   `scale = 1.0` → original power model (no change)
-   `scale = 0.5` → halve all alphas (more upsets, more randomness)
-   `scale = 1.5` → increase all alphas (fewer upsets, closer to pick favorite)
-   `scale = 0.0` → every game is a coin flip

**Q5.** Pick **two** different scale values and generate 500 brackets with each. Fill in your chosen values below and run the code.

```{r q7}
# CHANGE THESE to your chosen values
scale_a <- # pick scale A
scale_b <- # pick scale B

scale_model <- function(model, s) {
  mod <- model
  mod$r1_alphas$alpha <- mod$r1_alphas$alpha * s
  mod$round_alphas$alpha_bar <- mod$round_alphas$alpha_bar * s
  mod
}

mat_a <- build_power_prob_matrix(scale_model(power_mod, scale_a))
mat_b <- build_power_prob_matrix(scale_model(power_mod, scale_b))

scores_a <- replicate(500, score_bracket_fast(gen_R64(mat_a), actual))
scores_b <- replicate(500, score_bracket_fast(gen_R64(mat_b), actual))

comparison <- tibble(
  score = c(pf_scores, r64_scores, scores_a, scores_b),
  method = rep(c("Pick Favorite", "Power (original)",
                 paste0("Scaled (", scale_a, ")"),
                 paste0("Scaled (", scale_b, ")")), each = 500)
)

comparison %>%
  group_by(method) %>%
  summarise(mean = mean(score), max = max(score), sd = sd(score))
```

**Q6.** Compare the four models you just looked at using any visualizations or summary statistics you see fit. Which alpha value performed the best? Why do you think this is?

```{r q8}
# your code here
```

*Your answer:*

## Part 5: Comparing Generators

So far we've only used the **R64 generator**, which simulates every game starting from Round 1. But the paper proposes other strategies that "anchor" at later rounds. They first sample which seeds reach the Final Four or Elite Eight using historical distributions, then fill in the rest.

The intuition: if your model gets an early upset wrong, that error cascades through the whole bracket. Anchoring at a later round sidesteps that problem.

Run the code below to compare three generators using the same power model probabilities:

-   **R64:** Simulate all 63 games from Round 1
-   **F4A:** Sample 4 Final Four seeds from a historical distribution, then simulate the last 3 games
-   **E8:** Sample 8 Elite Eight seeds, then simulate from there

```{r generators}
sp <- compute_sampling_params(games, setdiff(1985:2023, 2020))

set.seed(345)
gen_results <- bind_rows(
  evaluate_gen(function() gen_R64(prob_mat), "R64", 500, 5, actual),
  evaluate_gen(function() gen_F4A(prob_mat, sp$q_f4), "F4A (anchor F4)", 500, 5, actual),
  evaluate_gen(function() gen_E8(prob_mat, sp$q_e8), "E8 (anchor E8)", 500, 5, actual),
  evaluate_gen(function() gen_PF(), "Pick Favorite", 500, 5, actual)
)
```

**Q7.** Compare the max scores and mean scores across generators. Which generator produced the single highest-scoring bracket? Which had the highest average?

```{r q9}
gen_results %>% select(generator, max_score_mean, max_score_sd, mean_score)
```

*Your answer:*

**Q8.** Now combine Parts 4 and 5. Take your best alpha scaling value from Part 4 and pair it with the generator that performed best above. Does the combination do better than either change alone?

```{r q11}
# Use your best scale from Part 4
best_scale <- # fill in your value
best_mat <- build_power_prob_matrix(scale_model(power_mod, best_scale))

# Use the best generator from above (change gen_F4A/gen_E8 if needed)
combo_scores <- replicate(500, score_bracket_fast(gen_F4A(best_mat, sp$q_f4), actual))

tibble(
  method = c("Power (original R64)", "Your best combo"),
  mean = c(mean(r64_scores), mean(combo_scores)),
  max = c(max(r64_scores), max(combo_scores)),
  sd = c(sd(r64_scores), sd(combo_scores))
)
```

*Your answer:*

**Q9.** In 1–2 sentences: the paper finds that no single generator is best every year. Why might that be the case? (Think about what kind of tournament each generator is best suited for.)

*Your answer:*

**Q10.** The power model has two notable limitations. First, it reduces every team to its seed number. For example, a dominant 1-seed and a weak 1-seed get identical win probabilities. Second, it is entirely circular for Round 1: the model estimates alpha values from observed win rates, then uses those alphas to reconstruct probabilities that exactly match the original win rates. Basically, the model doesn't learn anything beyond what the raw frequencies already tell you.

Given these limitations, what approach would you take instead? What information beyond seed number would you incorporate, and how might it help the model generalize rather than just memorize historical frequencies?

*Your answer:*
